\plaintitle

### 2016-09-20 Paper/proceedings.tex
JH edits.

% If you want to use todo notes, marginpars etc. during creation of your draft document, you
% have to enable the "chi_draft" option for the document class. To do this, change the very first
% line to: "\documentclass[chi_draft]{sigchi}". You can then place todo notes by using the "\todo{...}"
% command. Make sure to disable the draft option again before submitting your final document.
\usepackage{todonotes}

@@

\begin{abstract}
Observing trends and predicting future values are common tasks for viewers of bivariate data visualizations. As many charts do not explicitly include trend lines or related statistical summaries, viewers often visually estimate trends directly from a plot. How reliable are the inferences viewers draw when performing such \emph{regression by eye}? Do particular visualization designs or data features bias trend perception? We present a series of crowdsourced experiments that assess the accuracy of trends estimated using regression by eye across a variety of bivariate visualizations, and examine potential sources of bias in these estimations. We find that viewers accurately estimate trends in many standard visualizations of bivariate data, but that both visual features (e.g., ``within the bar'' bias) and data features (e.g., the presence of outliers) can result in visual estimates that systematically diverge from standard least-squares regression models. In most cases regression by eye performs similarly to ordinary least squares regression, but there are edge cases where perceptual and statistical regression systematically diverge. These results suggest that, while regression by eye can be a reliable way of estimating trends, in some common cases these estimates systematically differ from statistical methods of regression.
\jeff{These last three sentences seem a bit overlapping / redundant. Plus we really seem to like the word ``systematically''.}
\end{abstract}

@@

\section{Introduction}

%motivation here is poor. needs another turn of the crank.

An oft-cited example of the power of data visualization is Anscombe's quartet~\cite{anscombe1973graphs}, a set of four bivariate datasets with nearly identical summary statistics, but with qualitatively different patterns when drawn as four scatter plots. \jeff{Add figure of Anscombe's quartet...} This example relies on the fact that people have the ability to \emph{perceptually} estimate statistical quantities of interest. Visualization users regularly perform statistical tasks, including model selection, identification of outliers, and estimation of summary statistics, entirely through visual inspection. \jeff{Citations to perceptual work touching on the previous points? E.g., scatter plot means, color weaving, others?}

@@

In this work, we describe a series of crowdsourced experiments on the visual estimation of trends in common bivariate visualizations such as scatter plots, area charts, and line graphs. We present the results of three studies investigating estimation of trend slope, trend intercept, and the effect of outliers. We find that, while in most cases viewers accurately estimate trends, area charts introduce systematic underestimation of trend intercept, and outliers introduce ambiguity in how trends are estimated. \jeff{This is itself a bit ambiguous. What do you mean by ``introduce ambiguity''? Can you be more precise?} These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments. \jeff{Obligatory design implications remark here?}

@@

Our task of trend estimation combines elements of both prediction and correlation estimation tasks. As with prediction, there is not necessarily an unambiguous correct estimation of trend (different modeling and regression methods can produce different trend lines). As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.~\cite{harrison2014ranking}) likely relies on a set of key visual proxies. \jeff{What do you mean by ``key visual proxies''? Be clearer so that readers can follow along.} And, as with both tasks, we believe that trend estimation can be biased through conscious or unconscious design choices.
\jeff{Such as? The related work could end on a stronger note if we put forward some concrete hypotheses and note that prior work has not addressed them...}

@@

\subsection{Experimental Interface}

In each experiment, we presented participants with visualizations of bivariate data and asked them to interactively adjust trend lines to best fit that data. Participants responded using a slider without tick marks, in order to limit anchoring effects~\cite{matejka2016effect}. Moving the slider adjusted a purple trend line by modifying its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). Fig. \ref{fig:experiment} shows an example experimental task. After moving the slider, participants had to confirm their choice of trend line. The primary dependent measure is accuracy (e.g., the difference between the subject-specified slope and the slope of the series if the residuals were removed). We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand. We avoided more expressive designs (such as freeform drawing or other elicitation techniques) in order to constrain the model space and to reduce the impact of factors such as accuracy of motor movements or heterogeneity of input devices.

@@

@@

For each point set, we created a set of residual values, sampled evenly from a discretized Gaussian. This set was permuted, and then applied to the points along with a target trend. As heteroskedasticity introduced through permutation could alter the actual trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend. We reused these residuals across all different trend types (linear, quadratic, or trigonometric). Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$. In all experiments, we desired control over the direction of the trend. For linear fits, this is the slope of the trend line. For quadratic fits, this was the curvature, as controlled by the coefficient of the second degree term. For trigonometric fits, this was the amplitude of a cosine function (with negative amplitude corresponding to negative slopes, and positive to positive slopes). These terms create similar relationships, such that a value of $0$ is the line $f(x)=0.5$, a value of $1$ goes from $f(0)=0$ to $f(1)=1$, and $-1$ has the inverse relationship.

@@

We designed our first experiment to examine how accurate participants were at estimating the magnitude (slope, amplitude, or curvature) of trends in bivariate visualizations. We examined a variety of common bivariate visualizations \jeff{Such as?}. In addition to linear trends, we examined more complex relationships such as quadratic and trigonometric functions.

We presented participants with a series of bivariate visualizations, who adjusted a slider to fit the perceived trend in the points. The visualizations were one of a scatter plot, line graph, or area chart (with the filled area below the line). The slider parameterized one of three types of trend (linear, quadratic, or trigonometric). For each stimulus, participants adjusted a slider that controlled the slope of a rendered trend line (or, in the case of the quadratic trends, the curvature, or the trigonometric, the positive/negative amplitude). \jeff{To many parentheticals...}

Participants saw one of each combination of 3 chart types (scatter plot, line chart, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, and 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli. We also included an additional 4 validation stimuli otherwise excluded from analysis. The type of trend (linear, quadratic, or trigonometric) was a random factor, with 32 stimuli of each factor level. \jeff{Was this factor random or consistent across subjects?}

@@

\subsection{Results}
We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on unsigned error \jeff{absolute error?} in estimation of trend lines. We included participant ids and the actual slope of the trend line as covariates.

Our results support our first hypothesis: \textbf{larger residuals reduce accuracy at regression by eye}.  We observed a significant main effect for the bandwidth of the Gaussian used to generate residuals ($F(1,4554)=950$, \jeff{p-value?}). The trimmed mean \jeff{trimmed how?} of absolute error increased monotonically with this bandwidth, from $7\%$ of the actual slope of the trend line when the bandwidth was $0.05$, to $32\%$ when the bandwidth was $0.20$. Figure \ref{fig:sigma} illustrates these results.

Our results fail to support our second hypothesis: \textbf{there was no statistically significant difference in estimation accuracy among linear, quadratic, or trigonometric trends}. Fit type had only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and a post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise-interactions. The trimmed mean of absolute estimation error of linear fits was $17\%$ of the actual slope, in the middle of the $16\%$ error for quadratic fits, and $18\%$ for trignometric fits, providing evidence that the relative unfamiliarity of non-linear trends does not have a notable impact on performance.

Our results support our third hypothesis: \textbf{there was no statistically significant bias in estimations}. Participants saw an equal number of stimuli with positive and negative trend lines. If estimates of these trends were unbiased, we would expect the average signed error to be close to zero. The average signed error was $0.0008$, far less than the fidelity of the slider used to input guesses ($\Delta=0.01$). A Student's T-test failed to support the hypothesis that $mu_{error}\ne0$ ($p=0.70$) \jeff{Test statistic and degrees of freedom?}.

@@

@@

We had one hypothesis for the second experiment:
\begin{enumerate}
	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line charts and scatter plots.
\end{enumerate}

@@

Our results support our first hypothesis: \textbf{participants systematically overestimated the intercept of trends in area charts, but not in scatter plots or line graphs}. We observed a significant main effect of graph type on signed error ($F(2,2431)=27$, $p<0.001$). A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatter plots and line graphs. The trimmed mean \jeff{trimmed how?} of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to a trimmed mean of $0.002$ for the other two conditions. This underestimation corresponded to unsigned errors more than twice as large in area charts (trimmed mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for line graphs and scatter plots). Figure \ref{fig:outlier} illustrates this result.

@@

\outlierFig
OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit. Extreme outliers violate this assumption, and can result in trend lines that are substantially different from those produced by more robust methods. Visual inspection can accurately identify certain classes of outliers \cite{albers2014task}. Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends. However, cognitive biases such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in their estimations \jeff{Need citations in this sentence?}. These biases can impact forecasting tasks~\cite{campbell2009anchoring, ji2001culture}. That is, regression by eye may not uniformly weight outliers, depending on their position in the plot.

@@

We had three hypotheses for the third experiment:
\begin{enumerate}
	\item \textbf{Participant estimation would be closer to a robust trend line ignoring outliers, than the non-robust OLS trend line.} That is, we assumed that in general, participants would ignore or downweight outliers when performing regression by eye.
	\item \textbf{As the number of outliers increased, estimations would be closer to the non-robust OLS fit.} We speculated on the existence of a ``tipping point'' of outlier density, beyond which participants would avoid robust fits, and interpolate between the robust and OLS fits.
	\item \textbf{Outliers at the end of the chart would result in estimations closer to the OLS fit than outliers in other locations on the plot.} Prior work on cognitive biases in forecasting tasks suggests that viewers may give more credence to more recent outliers (as indicative as new anchors points, or of an emerging ``streak'') \jeff{cite?}. We therefore believed that these points would be more heavily weighted, and the resulting estimation closer to the outlier-sensitive OLS line, than outliers ``earlier'' in the plot.
\end{enumerate}

@@

Our results support our first hypothesis. \textbf{Participant estimates were closer to the robust trend line than the trend line that included outliers}. Excluding conditions with zero outliers (and so the two trend lines would be identical), the trimmed mean \jeff{trimmed how?} of unsigned error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to the robust trend line ($0.08$). A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($p<0.001$) \jeff{test statistic, dof}. However, a Student's T-test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($\mu=0.07$ vs. $\mu=0.08$, $p<0.001$). This indicates that participants are not entirely ignoring outliers, although the small effect size suggestes that participants are giving outliers significantly less weight than OLS regression.

\jeff{Whither the other hypotheses?}

%There is a confound between the number of outliers and the OLS trend line: since all outliers had approximately the same magnitude, more outliers create more extreme slopes with respect to the robust trend line, and so (if a participant were strictly ignoring outliers), larger differences in unsigned error. Similarly, outliers at the ends of the data series will adjust the slope of the OLS line more than outliers in the middle. We therefore measured performance as the unsigned error as a percentage of the difference between the slopes of the two types of fit. We performed a two-way ANOVA on the effect of outlier number and location on this percentage error measure. Participant id was included as a random variable.

%Our results fail to support our second hypothesis. \emph{Increasing number of outliers does not increase fidelity to the OLS trend line}. A one-way ANCOVA on the effect of outlier numbers greater than 0 on this percentage error showed that outliers were a significant main effect ($F(1,3438)=59$, $p<0.001$). However, the percentage error \emph{increases} as the number of outliers increase. A post-hoc Tukey's HSD shows that the percentage errors for 10 and 15 outliers are similar (trimmed means are $111\%$ and $112\%$ respectively), they are significantly different from the stimuli with only 5 outliers (trimmed mean of $97\%$).

@@

In many cases, designers do not explicitly encode regression information. In other cases, viewers may not have the statistical or graphical expertise to interpret such information, even when it is supplied. Yet, our results point to regression by eye as a robust and accurate method for estimating trends in bivariate data. Participants from a variety of backgrounds and levels of self-reported graphical and statistical expertise were capable of estimating both the slope and intercept of both linear and non-linear trends. That is, viewers of visualization are largely trustworthy when estimating the relationship between two variables in a plot.
\jeff{This point was not so clear in the results discussion of experiment 1. We should discuss (and visualize) the range of accuracies observed there.}

@@

Likewise, viewers do not give the same weight to outliers as OLS regression, but they do not ignore them either. For noisy data, this robustness may be beneficial: people can be relied on to perform filtering operations without explicit guidance. However, in other cases, this insensitivity may result in visual estimations at odds with the statistical modeling underpinning other areas of analysis of a particular dataset \jeff{What are you trying to say here?}. The emergence of new modes and the shifting of means may not result in a subsequent shift in beliefs about trends in data. \jeff{???}

@@

Our future work is focused in three areas. First, we wish to examine the impact of annotations relevant to regression (such as confidence bands, error bars, and curve boxplots~\cite{mirzargar2014curve}) on regression by eye. Can sufficient information promote caution in judgments of trend, or do initial visual estimations have too much weight to be overcome? Second, we wish to examine the impact of different rhetorical framings and presentations on regression by eye. Language from semiotics and rhetoric can provide testable structures for how visualizations are consumed~\cite{hullman2011visualization}: these framings (and so associated predispositions or biases about the data) could similarly impact statistical judgments. \jeff{Can you clarify this last part? And/or provide a concrete example?} Lastly, we wish to examine techniques for overcoming bias in regression by eye and similar visual estimations of statistical quantities. Cognitive and perceptual biases are difficult to overcome, and may require exploration of novel forms of visualization~\cite{micallef2012assessing}.

@@

In this paper, we examine the ability of viewers of visualizations to perform \emph{regression by eye}\,---\,the visual estimation of trends in bivariate visualizations. We show that even viewers without statistical training can reliably estimate both linear and non-linear trends in visualizations such as scatter plots and line graphs. However, area charts are subject to a ``within the bar'' bias, leading to estimated trends with lower intercepts than other bivariate visualizations. Regression by eye is also less sensitive to outliers than simple statistical regression techniques, resulting in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by simple statistical methods, can give equal weight all points in a dataset. \jeff{Run-on sentence. Also OLS does \emph{not} give equal weight to all points, such is the nature of \textbf{squared} error.}

>>>>

\usepackage{amsmath}
% If you want to use todo notes, marginpars etc. during creation of your draft document, you
% have to enable the "chi_draft" option for the document class. To do this, change the very first
% line to: "\documentclass[chi_draft]{sigchi}". You can then place todo notes by using the "\todo{...}"
% command. Make sure to disable the draft option again before submitting your final document.
\usepackage{todonotes}

@@

\begin{abstract}
Observing trends and predicting future values are common tasks for viewers of bivariate data visualizations. As many charts do not explicitly include trend lines or related statistical summaries, viewers often visually estimate trends directly from a plot. How reliable are the inferences viewers draw when performing such \emph{regression by eye}? Do particular visualization designs or data features bias trend perception? We present a series of crowdsourced experiments that assess the accuracy of trends estimated using regression by eye across a variety of bivariate visualizations, and examine potential sources of bias in these estimations. We find that viewers accurately estimate trends in many standard visualizations of bivariate data, but that both visual features (e.g., ``within the bar'' bias) and data features (e.g., the presence of outliers) can result in visual estimates that systematically diverge from standard least-squares regression models. In most cases regression by eye performs similarly to ordinary least squares regression, but there are edge cases where perceptual and statistical regression systematically diverge. These results suggest that, while regression by eye can be a reliable way of estimating trends, in some common cases these estimates systematically differ from statistical methods of regression.
\end{abstract}

@@

\section{Introduction}
\anscombeFig

An oft-cited example of the power of data visualization is Anscombe's quartet~\cite{anscombe1973graphs}, a set of four bivariate datasets with nearly identical summary statistics, but with qualitatively different patterns when drawn as four scatter plots (Fig. \ref{fig:anscombe}). This example relies on the fact that people have the ability to \emph{perceptually} estimate \emph{statistical} quantities of interest. Visualization users regularly perform statistical tasks, including model selection, identification of outliers, and estimation of summary statistics, entirely through visual inspection. Recent work has examined the accuracy of these visual inspections for means in scatter plots \cite{gleicher2013perception} and time series data \cite{correll2012comparing}, as well as speculated on the affordances of visualizations in general for supporting this sort of visual estimation of summary statistics \cite{szafir2016four}.

@@

In this work, we describe a series of crowdsourced experiments on the visual estimation of trends in common bivariate visualizations such as scatter plots, area charts, and line graphs. We present the results of three studies investigating estimation of trend slope, trend intercept, and the effect of outliers. We find that, while in most cases viewers accurately estimate trends, area charts introduce systematic underestimation of trend intercept, and that viewers give low weights to extreme values when estimating trends. These results suggest that there are several areas where human judgments diverge from the fitted models generated by techniques such as Ordinary Least Squares (OLS), and that the design of bivariate visualizations can introduce additional biases in these judgments. Designers should therefore

@@

Our task of trend estimation combines elements of both prediction and correlation estimation tasks. As with prediction, there is not necessarily an unambiguous correct estimation of trend (different modeling and regression methods can produce different trend lines). As with correlation, viewers must make holistic judgments about the dataset in a way that (as per Harrison et al.~\cite{harrison2014ranking}) likely relies on a set of visual proxies. That is, while viewers are unlikely to be directly estimation correlation \emph{per se}, they are estimating correlation through the perception of visual features such as the envelope of the points, point dispersion, or some other visual feature highly correlated with the statistic of interest.

Prior work in visualization has not directly addressed the capabilities of regression by eye, instead assuming as a given that visual estimates of trend are sufficiently accurate. Work in the perception of summary statistics lends credences to this assumption, but we believe that trend estimation can be biased through conscious or unconscious design choices. For instance, within-the-bar bias may result in underestimation of trends in bar and area charts, and factors that affect perception of correlation (such as aspect ratio and whitespace) may result in under- or overweighting of outliers.

@@

\subsection{Experimental Interface}
\typesFig

We investigated three types of common bivariate visualizations: scatter plots, line graphs, and area charts (see Fig. \ref{fig:types}). In each experiment, we presented participants with a visualization and asked them to interactively adjust trend lines to best fit the data.  Participants responded using a slider without tick marks, in order to limit anchoring effects~\cite{matejka2016effect}. Moving the slider adjusted a purple trend line by modifying its slope (in experiments 1 and 3) or its y-intercept (in experiment 2). Fig. \ref{fig:experiment} shows an example experimental task. After moving the slider, participants had to confirm their choice of trend line. The primary dependent measure is accuracy (e.g., the difference between the subject-specified slope and the slope of the series if the residuals were removed). We chose this design over a standard binary forced choice design for its greater expressiveness, as well as for the interactive feedback of adjusting the fit by hand. We avoided more expressive designs (such as freeform drawing or other elicitation techniques) in order to constrain the model space and to reduce the impact of factors such as accuracy of motor movements or heterogeneity of input devices.

@@

\sigmasFig

@@

For each point set, we created a set of residual values, sampled evenly from a discretized Gaussian. The bandwidth of this Gaussian controls both goodness of fit and dispersion of points (see Fig. \ref{fig:sigmas}). We permuted this set of residuals, and then applied them to points along a target trend. As heteroskedasticity introduced through permutation could alter the actual trend away from the target trend, we performed rejection sampling to ensure that the trend of the resulting points were within $10^{-7}$ of the target trend. We reused these residuals across all different trend types (linear, quadratic, or trigonometric). Except where noted, we selected trend lines that were centered in the image: that is, for a horizontal data extent of $[0,1]$, $f(0.5) = 0.5$. In all experiments, we desired control over the direction of the trend. For linear fits, this is the slope of the trend line. For quadratic fits, this was the curvature, as controlled by the coefficient of the second degree term. For trigonometric fits, this was the amplitude of a cosine function (with negative amplitude corresponding to negative slopes, and positive to positive slopes). These terms create similar relationships, such that a value of $0$ is the line $f(x)=0.5$, a value of $1$ goes from $f(0)=0$ to $f(1)=1$, and $-1$ has the inverse relationship.

@@

\expOnesigmasFig

We designed our first experiment to examine how accurate participants were at estimating the magnitude (slope, amplitude, or curvature) of trends in bivariate visualizations. We examined three types of bivariate visualizations: scatter plots, line graphs, and area chart (with the filled area below the line).In addition to linear trends, we examined more complex relationships such as quadratic and trigonometric functions.

We presented participants with a series of bivariate visualizations, who adjusted a slider to fit the perceived trend in the points. The slider parameterized one of three types of trend: linear, quadratic, or trigonometric. For each stimulus, participants adjusted a slider that controlled the slope of a rendered trend line. In the case of quadratic trends, this slope was the curvature; for trigonometric fits, the positive/negative amplitude.

Participants saw one of each combination of 3 chart types (scatter plot, line graph, or area chart), 8 possible slopes $\beta = \pm \{0.1,0.2,0.4,0.8\}$, and 4 bandwidths of Gaussian residuals $\sigma = \{0.05,0.1,0.15,0.2\}$, for a total of 96 stimuli. We also included an additional 4 validation stimuli otherwise excluded from analysis. The type of trend (linear, quadratic, or trigonometric) was a random factor, with 32 stimuli of each factor level randomly assigned.

@@

\subsection{Results}
We performed a three-way analysis of covariance (ANCOVA) of the effect of residual bandwidth, graph type and trend type on absolute error in estimation of trend lines. We included participant ids and the actual slope of the trend line as covariates.

Our results support our first hypothesis: \textbf{larger residuals reduce accuracy at regression by eye}.  We observed a significant main effect for the bandwidth of the Gaussian used to generate residuals ($F(1,4554)=950$, $p<0.001$). The interquartile mean of absolute error increased monotonically with this bandwidth, from $7\%$ of the actual slope of the trend line when the bandwidth was $0.05$, to $32\%$ when the bandwidth was $0.20$. Figure \ref{fig:exp1sigmas} illustrates these results.

Our results fail to support our second hypothesis: \textbf{there was no statistically significant difference in estimation accuracy among linear, quadratic, or trigonometric trends}. Fit type had only a marginal main effect ($F(2,4554)=2.60$, $p=0.074$), and a post-hoc tests (using Tukey's Honest Significant Difference) did not identify any significant pairwise-interactions. The interquartile mean of absolute  error of linear fits was $17\%$ of the actual slope, in the middle of the $16\%$ error for quadratic fits, and $18\%$ for trignometric fits, providing evidence that the relative unfamiliarity of non-linear trends does not have a notable impact on performance.

Our results support our third hypothesis: \textbf{there was no statistically significant bias in estimations}. Participants saw an equal number of stimuli with positive and negative trend lines. If estimates of these trends were unbiased, we would expect the average signed error to be close to zero. The average signed error was $0.0008$, far less than the fidelity of the slider used to input guesses ($\Delta=0.01$). A Student's T-test failed to support the hypothesis that $mu_{error}\ne0$ ($t(4590) = 0.39$, $p=0.70$).

In general, accuracy at estimation of slope was high, an interquartile mean of $0.06$ for slope values that ranged from $\{-1,1\}$, corresponding to an accuracy of $17\%$ of the actual slope.

@@

\expTwoTypesFig

@@

We had one hypothesis for the second experiment:
\begin{enumerate}
	\item \textbf{Area charts would be subject to within the bar bias.} That is, participants would estimate lower values of y-intercepts of trends than with line graphs and scatter plots.
\end{enumerate}

@@

Our results support our first hypothesis: \textbf{participants systematically underestimated the intercept of trends in area charts, but not in scatter plots or line graphs}. We observed a significant main effect of graph type on signed error ($F(2,2431)=27$, $p<0.001$). A post-hoc Tukey's HSD confirmed significant differences in error between the area chart and the other two chart types, but not between scatter plots and line graphs. The interquartile mean of the signed error of estimations made with area charts was an underestimation of $-0.02$, compared to an interquartile mean of $0.002$ for the other two conditions. This underestimation corresponds to unsigned errors more than twice as large in area charts (interquartile mean of $0.04$ for intercepts that ranged from $\{-0.25,0.25\}$, compared to $0.02$ for line graphs and scatter plots). Figure \ref{fig:exp2types} illustrates this result.

@@

\outlierFig
OLS regression operates under the assumption that there is a unimodal, symmetric distribution of residuals surrounding the line of best fit. Extreme outliers violate this assumption, and can result in trend lines that are substantially different from those produced by more robust methods. Visual inspection can accurately identify certain classes of outliers \cite{albers2014task}. Therefore, regression by eye may afford the estimation of both outlier-robust and outlier-sensitive trends. However, cognitive biases~\cite{tversky1975judgment} such as anchoring (the tendency to overweight the first set of information), availability (the tendency to overweight more recent or extreme information), and the hot-hand fallacy (the tendency to assume that trends of high or low values with continue) can impact how viewers categorize and utilize outliers in prediction tasks ~\cite{campbell2009anchoring, ji2001culture}. That is, regression by eye may not uniformly weight outliers, depending on their position in the plot.

@@

We had three hypotheses for the third experiment:
\begin{enumerate}
	\item \textbf{Participant estimation would be closer to a robust trend line ignoring outliers, than the non-robust OLS trend line.} That is, we assumed that in general, participants would ignore or downweight outliers when performing regression by eye.
	\item \textbf{As the number of outliers increased, estimations would be closer to the non-robust OLS fit.} We speculated on the existence of a ``tipping point'' of outlier density, beyond which participants would avoid robust fits, and interpolate between the robust and OLS fits.
	\item \textbf{Outliers at the end of the chart would result in estimations closer to the OLS fit than outliers in other locations on the plot.} Prior work on cognitive biases in predictions suggests that viewers may give more credence to more recent outliers (as indicative as new anchors points, or of an emerging ``streak'')~\cite{croson2005gambler}. We therefore believed that these points would be more heavily weighted, and the resulting estimation closer to the outlier-sensitive OLS line, than outliers ``earlier'' in the plot.
\end{enumerate}

@@

Our results support our first hypothesis. \textbf{Participant estimates were closer to the robust trend line than the trend line that included outliers}. Excluding conditions with zero outliers (and so the two trend lines would be identical), the interquartile mean of absolute error was over $4$ times higher when calculated as a comparison to the OLS trend line ($0.36$) as opposed to the robust trend line ($0.08$). A Student's T-test confirmed that the error as defined by the OLS line was significantly higher than the robust line ($t(6670)= 43$, $p<0.001$). However, a Welch Two Sample T-test found a significant difference in unsigned error when compared to the robust trend line between estimates where no outliers were present, and those with any amount of outliers ($\mu=0.07$ vs. $\mu=0.08$, $t(3808) = 9.8$, $p<0.001$). This indicates that participants are not entirely ignoring outliers, although the small effect size suggests that participants are giving outliers significantly less weight than OLS regression.

Our results fail to support our second hypothesis. \textbf{While participant estimates diverge from the robust line as the number of outliers increases, the OLS trend line diverges faster than participant estimates}. There was a significant positive interaction between outlier count and absolute error with respect to the robust trend line ($R^2 = 0.02$,$F(1,4591)=110$, $p<0.001$). However, there was also a significant positive interaction between outlier count and absolute error with respect to the OLS line  ($R^2 = 0.38$,$F(1,4591)=2800$, $p<0.001$). Additionally, while the range of errors with respect to the robust line remained small (interquartile means of $0.07$ with 0 outliers, monotonically increasing to $0.10$ with 15 outliers), absolute error with respect to the OLS line increased almost 7-fold, from $0.07$ with 0 outliers, monotonically to $0.54$ with 15 outliers. These results indicate that while increased numbers of extreme values may result in participants increasing the weight of these outliers, this increase in weight is slower than the sensitivity of OLS to outliers.

Our results also fail to support our third hypothesis, and seem to support its inverse: \textbf{There was no significant impact of outlier location on performance}. There is a confound between outlier location and OLS line: outliers on the ends of the data ranges create larger shifts in slope than outliers in the center, which may only create shifts in intercept. To evaluate our third hypothesis, we therefore defined a parameterized performance function $\frac{\text{Estimated Slope}-\text{Robust Slope}}{\text{OLS Slope} - \text{Robust Slope}}$ that contextualizes a participant estimate as an interpolation between the robust slope (value of $0$) and the OLS slope (value of $1$). A one way ANCOVA of this interpolation value as an effect of outlier location, with outlier count as a covariate, and participant ID as a random factor, found no significant effect ($F(2,3441)=0.63$, $p=0.53$).

%Contrary to our hypothesis, outliers in the middle location resulted in guesses that were closer to the trend line (interquartile mean of interpolation value of $0.66$, compared to $0.24$ for outliers at the beginning location, and $0.32$ for outliers in the final location).

@@

In many cases, designers do not explicitly encode regression information. In other cases, viewers may not have the statistical or graphical expertise to interpret such information, even when it is supplied. Yet, our results point to regression by eye as a robust and accurate method for estimating trends in bivariate data. Participants from a variety of backgrounds and levels of self-reported graphical and statistical expertise were capable of estimating both the slope and intercept of both linear and non-linear trends. That is, viewers of visualization are largely trustworthy when estimating the relationship between two variables in a plot.

@@

Likewise, viewers do not give the same weight to outliers as OLS regression, but they do not ignore them either. For noisy data, this robustness may be beneficial: people can be relied on to perform filtering operations without explicit guidance. In other cases, this insensitivity may be undesirable. For instance, more complicated automatic analytical processes may use OLS regression or other outlier sensitive techniques as part of their calculations, resulting in a disconnect between human and statistical conclusions at later stages of the sensemaking process. Similarly, the human tendency to discount or downweight outliers may result in slow adaptation to new data that does not fit into existing patterns.

@@

Our future work is focused in three areas. First, we wish to examine the impact of annotations relevant to regression (such as confidence bands, error bars, and curve boxplots~\cite{mirzargar2014curve}) on regression by eye. Can sufficient information promote caution in judgments of trend, or do initial visual estimations have too much weight to be overcome? Second, we wish to examine the impact of different rhetorical framings and presentations on regression by eye. Language from semiotics and rhetoric can provide testable structures for how visualizations are consumed~\cite{hullman2011visualization}: these framings (and so associated predispositions or biases about the data) could similarly impact statistical judgments. For instance, outliers from a source that is perceived as credible could be weighted higher than outliers with more suspect provenance. Likewise, preconceptions about the volatility of data from certain domains may color how experts estimate trends. Lastly, we wish to examine techniques for overcoming bias in regression by eye and similar visual estimations of statistical quantities. Cognitive and perceptual biases are difficult to overcome, and may require exploration of novel forms of visualization~\cite{micallef2012assessing}.

@@

In this paper, we examine the ability of viewers of visualizations to perform \emph{regression by eye}\,---\,the visual estimation of trends in bivariate visualizations. We show that even viewers without statistical training can reliably estimate both linear and non-linear trends in visualizations such as scatter plots and line graphs. However, area charts are subject to a ``within the bar'' bias, leading to estimated trends with lower intercepts than other bivariate visualizations. Regression by eye is also less sensitive to outliers than simple statistical regression techniques. This decreased sensitivity results in a divergence between trends estimated by viewers, which do not give much weight to extreme outliers, and those trends calculated by simple statistical methods, which can be heavily influenced by a few extreme values.
